Generalized Plackett-Luce Likelihoods

### 2022-08-12 inst/jss_hyper3.Rnw
slight rewording
cite some relevant work

\section{introduction}

The \code{hyper2} package~\citep{hankin2009,hankin2017} furnishes
computational support for generalized
Plackett-Luce~\citep{plackett1975} likelihood functions.  Our
preferred interpretation is a race (as in track and field atheletics):
Given six competitors $1-6$, we ascribe to them nonnegative strengths
$p_1,\ldots, p_6$; the probability that $i$ beats $j$ is
$p_i/(p_i+p_j)$.  It is conventional to normalise so that the total
strength is unity, and to identify a competitor with his strength.
Given an order statistic, say $p_1\succ p_2\succ p_3\succ p_4$, the
Plackett-Luce likelihood function would be

@@

The \code{hyper2} package generalizes this likelihood function to
functions of ${\mathbf p}=(p_1,\ldots,p_n)$ with

@@

\noindent where~$\mathcal{O}$ is a set of observations and~$s$ a
subset of~$\left[n\right]=\left\{1,2,\ldots,n\right\}$; numbers~$n_s$
are integers which may be positive or negative.  It becomes possible
to analyse a wider range of observations such as incomplete order
statistics or teams; and the introduction of reified entities allows
one to consider draws~\citep{hankin2009}, noncompetitive tactics such
as collusion~\citep{hankin2020}, and the phenomenon of team cohesion
wherein the team becomes stronger than the sum of its parts.

However, consider a race in which there are six runners 1-6 but we
happen to know that three of the runners (1,2,3) are clones of
strength $p_a$, two of the runners (4,5) have strength $p_b$, and the
final runner (6) is of strength $p_c$.  We normalise so
$p_a+p_b+p_c=1$.  The runners race and the finishing order is:

@@

In Formula 1 motor racing, the constructors' championship takes place
between {\em manufacturers} of racing cars (compare the drivers'
championship, which is between drivers).  Here, we consider the
constructor to be the object of inference.  Each constructor typically
fields two cars, each of which separately accumulates ranking-based
points at each venue.  Here we use generalized Plackett-Luce to assess
the contructors' performance.  The following table, included in the
\code{hyper2} package as a dataset, shows rankings for the first 9
venues of the 2021 season:

>>>>

\section{Introduction}

The \code{hyper2} package~\citep{hankin2009,hankin2017} furnishes
computational support for generalized
Plackett-Luce~\citep{plackett1975} likelihood functions.  Our
preferred interpretation is a race (as in track and field athletics):
Given six competitors $1-6$, we ascribe to them nonnegative strengths
$p_1,\ldots, p_6$; the probability that $i$ beats $j$ is
$p_i/(p_i+p_j)$.  It is conventional to normalise so that the total
strength is unity, and to identify a competitor with his strength.
Given an order statistic, say $p_1\succ p_2\succ p_3\succ p_4$, the
Plackett-Luce likelihood function would be

@@

\citet{mollica2014} call this a forward ranking process on the grounds
that the best (preferred; fastest; chosen) entities are identified in
the same sequence as their rank.  The \code{hyper2} package
generalizes this likelihood function to functions of ${\mathbf
  p}=(p_1,\ldots,p_n)$ with

@@

\noindent where~$\mathcal{O}$ is a set of observations and~$s$ a
subset of~$\left[n\right]=\left\{1,2,\ldots,n\right\}$; numbers~$n_s$
are integers which may be positive or negative.  It becomes possible
to analyse a wider range of likelihood functions than standard
Plackett-Luce~\citep{turner2020}.  For example, results involving
incomplete order statistics or teams are tractable. Further, the
introduction of reified entities allows one to consider
draws~\citep{hankin2009}, noncompetitive tactics such as
collusion~\citep{hankin2020}, and the phenomenon of team cohesion
wherein the team becomes stronger than the sum of its parts.

Here I present a different generalization.  Consider a race in which
there are six runners 1-6 but we happen to know that three of the
runners (1,2,3) are clones of strength $p_a$, two of the runners (4,5)
have strength $p_b$, and the final runner (6) is of strength $p_c$.
We normalise so $p_a+p_b+p_c=1$.  The runners race and the finishing
order is:

@@

In Formula 1 motor racing, the constructors' championship takes place
between {\em manufacturers} of racing cars (compare the drivers'
championship, which is between drivers).  In this analysis, the
constructor is the object of inference.  Each constructor typically
fields two cars, each of which separately accumulates ranking-based
points at each venue.  Here we use generalized Plackett-Luce to assess
the contructors' performance.  The following table, included in the
\code{hyper2} package as a dataset, shows rankings for the first 9
venues of the 2021 season:
